---
title: Rounding to Decimal Digits in Binary
author: Martin Mächler
address: ETH Zurich and R Core Team
date: 'July 4, 2020; rendered on `r Sys.Date()`'
output:
  html_vignette:
    css: style.css
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Rounding to Decimal Digits in Binary}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{round}
---

## Intro

In R (and its predecessors S and S+), we have always known and often used `round(x, digits)` to round a numeric
(or complex) vector of numbers `x` to `digits` decimals after the (decimal) point.
However, not many R users (nor scientists for that matter) have been aware of the fact that such
rounding is not trivial because our computers use binary (_base_ __2__) arithmetic and we are rounding to
decimal digits, aka decimals, i.e., in _base_ __10__.

On the topic of floating point computation, we have had the _most_ frequently asked question (FAQ) about R,
the infamous [R FAQ 7.31][],
and in 2017, Romain François even created an R package [seven31][] (not on CRAN) to help useRs exploring what we say in the FAQ.

<!-- ~/R/MM/NUMERICS/round-even-odd.R  -->
Recently, there has been an official R bug report (on R's bugzilla <bugs.r-project.org>), [`PR#17668`][]
with summary title __Artificial numerical error in `round()` causing round to even to fail__.

Adam Wheeler started with a shorter version (just using digits = 1,2,..,8) of the following examples and his own remarks about correctness:
```{r}
op <- options(digits = 15)
##' maybe add to package -- it's really useful here :
formatF <- function(x, ...) noquote(format(x, scientific=FALSE, drop0trailing=TRUE, ...))
formatF(rbind(           # R <= 3.6.x :
round(55.5, 0),          # 56 <- correct
round(55.55, 1),         # 55.5
round(55.555, 2),        # 55.55
round(55.5555, 3),       # 55.556 <- correct
round(55.55555, 4),      # 55.5555
round(55.555555, 5),     # 55.55555
round(55.5555555, 6),    # 55.555555
round(55.55555555, 7),   # 55.5555556 <- correct
round(55.555555555, 8),  # 55.55555555
round(55.5555555555, 9), # 55.555555555
round(55.55555555555,10),# 55.5555555556 <- correct
round(55.555555555555,11)# 55.55555555556 <- correct
))
```
(Note that we eventually will *not agree* on the above `correct` judgements,
 as the matter is more subtle than one might think at first)

Whereas the exact result of the R code above currently depends on your version of R, our `round` package's `roundX(x, dig, version = "r1.C")` now provides these, using the same C source code as R 3.6.2
(Note we adopt the convention to use `"r<n>.C"`, ending
 in `.C` for round()ing versions where R calls package C code):
```{r}
require(round)
formatF(rbind(
  roundX(55.5, 0, "r1.C"),
  roundX(55.55, 1, "r1.C"),
  roundX(55.555, 2, "r1.C"),
  roundX(55.5555, 3, "r1.C"),
  roundX(55.55555, 4, "r1.C"),
  roundX(55.555555, 5, "r1.C"),
  roundX(55.5555555, 6, "r1.C"),
  roundX(55.55555555, 7, "r1.C"),
  roundX(55.555555555, 8, "r1.C"),
  roundX(55.5555555555, 9, "r1.C"),
  roundX(55.55555555555,10, "r1.C"),
  roundX(55.555555555555,11, "r1.C")))
```

Adam (see above) used his own C code to see what happens in R's C code for `round()` and
 proposed to simplify the C code, *not* doing offset calculations (which substract the integer part `intx`, round and re-add `intx`), as now available with `roundX(*, "r0.C")`.
This has started an _"investigative"_ story which got quite a bit longer than anticipated.

## The Easy Problem "in Theory"

### Rounding to Integers

Well, rounding is supposedly quite simple and most of us learned a version of it in our
first years of school when learning the (decimal) numbers, calculating and then simple decimal fractions.
What most pupils learn (first) is round _up_, and round _down_ (to an _integer_), and then end with the "best"
__rounding to nearest__ i.e., rounding to the nearest _integer_, notably using _"round half up"_, which mathematically is
simply computing $y = r_0(x) := \left\lfloor x + 0.5 \right\rfloor$, see e.g., [Wikipedia, Rounding][],
or when taking _negative_ numbers into account and ensuring _symmetry_,
$y =  \mathrm{sgn}(x) \left\lfloor \left| x \right| + 0.5 \right\rfloor
   = -\mathrm{sgn}(x) \left\lceil -\left| x \right| - 0.5 \right\rceil$, where the
     _"floor"_ $\lfloor x \rfloor$
 and _"ceiling"_  $\lceil x \rceil$ operators are defined (customary in mathematics) as

$$ \lfloor x \rfloor := \max_{n \in \mathbb{Z}} \{n \le x\}, $$  and
$$ \lceil  x \rceil  := \min_{n \in \mathbb{Z}} \{n \ge x\}. $$

Whereas these (_rounding to nearest_  and _"round half up"_) rules, also called _"commercial rounding"_, are widely taught and used,
they lead to a small bias e.g., when numbers have all the same sign, and for this reason, for computing,
__round_half_to_even__ has been proposed and adopted as default rounding mode in
the IEEE 754 floating-point standard and it corresponds to C and C++ math library function `nearbyint()`,
and has been called _convergent rounding_, _statistician's rounding_, _bankers' rounding_ (and more),
see [Wikipedia, Round half to even][].
It is also what you get with R's `round(x)`, as `round()`'s second argument has default 0:
```{r, str-round}
str(round)
```
So, for now, let's set and use

$$ round(x) := r(x) := nearbyint(x)
$$

Note that all this rounding _to integer_ is defined _independently_ of the base of
your number representation system (such as _"decimal"_ or _"binary"_ ).

### Rounding to non-zero Digits

Things change slightly, when rounding to a non-zero number of decimal places, i.e., _digits_.
In theory, i.e., if computers would compute mathematically perfectly accurately, also this has a simple solution:

$$ round(x, d) := r(x \cdot 10^d) / 10^d
$$

for all integer $d \in \mathbb{Z}$ (i.e., including negative `digits` $d$).

All the practical problems stem from the fact that on a (binary arithmetic) computer,
the number $x/10$ cannot be represented exactly in binary unless $x$ is an integer multiple of 5.
Indeed, the simple fraction  1/5  is an infinite length binary fraction:
$$ \frac 1 5 = \frac{1 \cdot \frac{3}{16}}{5 \cdot \frac{3}{16}} = \frac{\frac{3}{16}}{1  - \frac{1}{16}}
 = \frac{3}{16} \frac{1}{1  - \frac{1}{16}}
 = \left(\frac{1}{8} + \frac{1}{16}\right) \cdot \left(1 + \frac{1}{16} + \frac{1}{16^2} + \ldots\right) \\
 = (0.001100110011001100110011\ldots\ldots)_2 .
$$


## Versions of round()ing - The Story

At first, [R bugzilla, comment #6][],
I've committed a version of Adam's proposal to R-devel ([R svn r77609][], on 2019-12-21, [^1]) but found that the simplification improved the above examples in that it always rounded to even, but it clearly broke cases that were working correctly in R 3.6.x.  That version is available with our
`roundX(*, version = "r0.C")`  (Version `0` as it is even simpler than version `1`).

One CRAN package had relied on  `round(x, digits = .Machine$integer.max)` to return integers unchanged, but
```{r}
roundX(c(-2,2), digits = .Machine$integer.max, version = "r0.C")
```
gave non-sense,
and there were less extreme cases of relatively large `digits` which had
stopped working with "r0".  See the two `roundX()` versions via simple wrapper `roundAll()`:
```{r, integer-x-large-digs}
i <- c(-1,1)* 2^(33:16)
stopifnot(i == floor(i)) # are integer valued
roundAll(i, digits = 300, versions = c("r0.C", "r1.C"))
```

Looking at these, I also found that internally, R's `round()` had effectively worked as if
`digits <- pmin(308, digits)`, i.e., truncated digits larger than 308.
This is clearly not
good enough for very small numbers (in absolute value),
```{r, small-x-large-digs}
e <- 5.555555555555555555555e-308
d <- 312:305 ; names(d) <- paste0("d=", d)
roundAll(e, d, versions = c("r0.C", "r1.C", "r2.C"))
```

As I was embarrassed to have blundered, I've worked and committed what now corresponds to
`roundX(*, version = "r2.C")` to R-devel ([R svn r77618][], on 2019-12-24, 16:11).

Also, Jeroen Ooms, maintainer of CRAN package `jsonlite`, contacted the CRAN team and me about the change in R-devel which broke one regression test of that package, and on Dec 27, he noticed that
R 3.6.2's version of `round()` was seemingly compatible [^2] with the (C library dependent)
versions of `sprintf()` and also with R's `format()`
whereas the R-devel versions where not, for his example:
(Note: `format(x, digits = d)` uses __significant__ digits `d`, not digits after the decimal point as `round()` does!)

```{r, JO-ex-sprintf}
x <- 9.18665
format(x, digits = 5) #  "9.1867"
sprintf("%.4f", x)    #  "9.1867"
## but -- showing now
roundAll(x, 4)

## but note that
print(x, digits=18)
```

which (typically) shows `9.1866500000000002`, i.e., a number closer to 9.1867 than to 9.1866, and so really should be rounded up, not down.
However, that is partly wrong: Whereas it is true that it's closer to 9.1867 than to 9.1866, one must be aware that these two decimal numbers are __neither__ exactly representable in binary double precision, and a further careful look shows actually the double precision version of these rounded numbers do have the exact same distance to `x` and that the main principle *round to nearest* here gives a tie:

```{r, JO-ex-diff}
print(rbind(9.1866, 9.18665, 9.1867), digits=18)
(dx <- c(9.1866, 9.1867) - x) # [1] -4.99999999998835e-05  4.99999999998835e-05
diff(abs(dx)) # is zero !
options(digits=7) # revert to (typical) default
```

and because of the tie, the *round to even* rule must apply which means rounding *down* to 9.1866, and so both libc's printf and hence R's `sprintf()` are as wrong as R 3.6.x has been, and indeed all our `roundX()` versions apart from `"sprintf"` and the '"r1*"' (previous R) ones, do round down:

```{r, roundAll-x}
roundAll(x, 4)
```

Finally, I think we've seen the light and on one hand recalled what we have known for long (but most R users will not be aware of at all)

> 1. Almost all finite decimal fractions are __not__ (exactly) representable as binary
> double precision numbers,

and consequently,

> 2. *round to nearest* applies much more often _directly_ rather than via the
> tie breaking rule *round to even* even for the case where the decimal fraction ends in a `5`.

and hence, a _"correct"_ implementation must really _measure, not guess_ which of the two possible decimals is closer to `x`.
This lead to our R level algorithm  `round_r3()`  which is the workhorse used by
`roundX(x,d, version = "r3")` :

```{r, round_r3-def}
round_r3
```

and its two C level versions `"r3.C"` (using `long double`) and `"r3d.C"` (_"d"_ for _"double"_, as it uses double precision only).

For R 4.0.0, this (equivalent of `"r3d.C")`, has been used for `round(x, digits)` now,
as does not use `long double` and is very close to the R level implementation `"r3"` (i.e., `round_r3()`) makes it potentially less platform dependent and easier to explain and document.

Lastly, note that the original set of examples is then treated differently from all previous proposals:

```{r, roundAll-55.5}
op <- options(digits = 15, width = 2*3*23)
formatF(rbind(
 d0 = roundAll(55.5, 0)
,d1 = roundAll(55.55, 1)
,d2 = roundAll(55.555, 2)
,d3 = roundAll(55.5555, 3)
,d4 = roundAll(55.55555, 4)
,d5 = roundAll(55.555555, 5)
,d6 = roundAll(55.5555555, 6)
,d7 = roundAll(55.55555555, 7)
,d8 = roundAll(55.555555555, 8)
,d9 = roundAll(55.5555555555, 9)
,d10= roundAll(55.55555555555,10))); options(op)
```

## Alternative Approaches

As I had asked for comments about my proposal(s), on the R-devel mailing lists,
Steven Dirkse (@ GAMS) mentioned he was not quite happy with the considerations above, and in the end (not on-list) summarized
his own rational approach in the following way -- which I like to present as it is coherent and simple, summarized as:

1. all double precision numbers are rationals mathematically, i.e., \(\in \mathbb{Q}\).
2. `round(x, n)` as a mathematical function is well defined unambigously on the rationals.
  In our notation above, simply \(round(x, n) := r(x \cdot 10^n) / 10^n\), where \(r(x) := round(x) = nearbyint(x)\),
  using _round to even_ if desired.

These two define _exact rounding_ and he would want R to use that.
Note for that, R would have to use exact arithmetic with rational numbers which has been available for years via the C library `GNU MP` aka [`GMP`] and in R via CRAN pkg [gmp][].

However, even if R would _"come with GMP"_ in the future (which is conceivable for other reasons),
we should note that this would still keep  `rx <- round(x, n)`  in R inaccurate in most cases, since `rx` is `numeric`, i.e., a double precision number, and almost all rational numbers are _not_ exactly representable as double.


[R FAQ 7.31]: https://cran.r-project.org/doc/FAQ/R-FAQ.html#Why-doesn_0027t-R-think-these-numbers-are-equal_003f

[seven31]: https://github.com/ThinkR-open/seven31 "seven31"

[`PR#17668`]: https://bugs.r-project.org/bugzilla/show_bug.cgi?id=17668

[Wikipedia, Rounding]: https://en.wikipedia.org/wiki/Rounding

[Wikipedia, Round half to even]: https://en.wikipedia.org/wiki/Rounding#Round_half_to_even

[R bugzilla, comment #6]: https://bugs.r-project.org/bugzilla/show_bug.cgi?id=17668#c6

[`GMP`]: https://gmplib.org/

[gmp]: https://CRAN.R-project.org/package=gmp

[^1]: using Winston Chang's semi-official mirror of R's official SVN repository at <https://svn.r-project.org/R/>
      (as `https://svn.r-project.org/R/trunk@77609` has not been enabled in its server)

[^2]: do note that `sprintf()` and R 3.6.x's version of `round()` are *not* at all equivalent, even if they are for Jeroen's example

[R svn r77609]: https://github.com/wch/r-source/commit/33d95f1
[R svn r77618]: https://github.com/wch/r-source/commit/7a55bbeb

### Session information
(Note half a dozen non-standard packages present only as dependences of `rmarkdown` we use for rendering this vignette)
```{r, echo=FALSE}
print(sessionInfo(), locale=FALSE)
```
